Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-01 paper.tex
supp tables started
placeholder figure
fill in a few things
paper updates

% local definitions
\newcommand{\comment}[1]{{\textcolor{red}{Comment: #1}}}

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
\caption{\textbf{The first figure}.
    It has a caption.
}
\label{fig:1}
\end{center}
\end{figure}

\subsection*{A Late Middle Stone Age common ancestry for contemporary human populations}

\subsection*{Evidence for deep population structure, but not archaic admixture within Africa}

@@

>>>>

% local definitions
\newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
\newcommand{\bmhcomment}[1]{{\textcolor{blue}{BMH: #1}}}
\newcommand{\aprcomment}[1]{{\textcolor{purple}{APR: #1}}}
\newcommand{\tdwcomment}[1]{{\textcolor{orange}{TDW: #1}}}

@@

\subsection*{A Late Middle Stone Age common ancestry for contemporary humans}

\subsection*{Deep population structure but not archaic admixture within Africa}

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
\caption{\textbf{The first figure}.
    The geographic and genetic diversity of populations across Africa.
}
\label{fig:1}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{} % \includegraphics{figures/fig2.png}}
\caption{\textbf{The second main figure}.
    A placeholder - best fit model(s).
}
\label{fig:2}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
\caption{\textbf{The third main figure}.
    A placeholder - validation (Relate and cSFS).
}
\label{fig:r3}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
\caption{\textbf{The fourth main figure}.
    A placeholder - predictions (FST and/or f4).
}
\label{fig:r4}
\end{center}
\end{figure}

### 2022-01-01 supplement.tex
paper updates

DNA samples were collected in South Africa (Richtersveld) with IRB approval
from Stanford University, Stellenbosch University (ZA) and SUNY Stony Brook.
\aprcomment{Does this need to be expanded on at all?}

@@

\subsection{Computing LD and diversity statistics}

@@

\subsection{Computing conditional SFS}

The conditional site frequency spectrum (or cSFS) is the distribution of allele
frequencies restricted to loci that satisfy a given condition. Specifically, we
consider the distribution of allele frequencies in present-day populations
conditioned on the Vindija Neanderthal carrying the derived allele relative to
the inferred ancestral allele. Ancestral alleles alleles were determined from a
6 primate alignment \aprcomment{cite}. This cSFS is expected to be close to
uniform under neutrality and a simple split model (with no subsequent
migration) between the ancestors of modern humans and Neanderthal
\citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
evidence for archaic introgression from a population whose split from modern
humans is at least as old as that of Neanderthal’s
\citet{Durvasula2020-td,Yang2012-ze}. Because we wanted to compare our
inferences (based on intergenic sites) to inferences from previous work (based
on whole-genome data), we computed the cSFS for both intergenic and all sites
genome-wide. Sites with no calls in the Vindija Neanderthal were excluded from
this analysis. \aprcomment{Figures XX.}

@@

@@

@@

@@

>>>>

DNA samples were collected from three Nama communities in the Richtersveld
region of South Africa, which borders southern Namibia, in 2012. A community
guide was present during each interview and facilitated consent in Afrikaans or
Nama. Written consent was recorded per our IRB protocol with human subjects
approval from Stanford University (Protocol \#13829), Stellenbosch University
(N11/07/210) and later maintained via SUNY Stony Brook (Protocol \#727494).
Saliva samples were extracted from Oragene [OGR-500] kits at Stellenbosch
University. Results stemming from genetic analyses have been communicated in
2015, 2019, 2021 via community presentations, a radio interview and to
representatives of the Richtersveld National Park.

@@

\subsection{Computing LD and diversity statistics}\label{sec:computing-stats}

@@

\subsection{Computing conditional SFS}\label{sec:computing-csfs}

The conditional site frequency spectrum (or cSFS) is the distribution of allele
frequencies restricted to loci that satisfy a given condition. Specifically, we
consider the distribution of allele frequencies in present-day populations
conditioned on the Vindija Neanderthal carrying the derived allele relative to
the inferred ancestral allele. Ancestral alleles alleles were determined from a
6 primate alignment \aprcomment{cite}. This cSFS is expected to be close to
uniform under neutrality and a simple split model (with no subsequent
migration) between the ancestors of modern humans and Neanderthal
\citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
evidence for archaic introgression from a population whose split from modern
humans is at least as old as that of the human-Neanderthal split
\citet{Durvasula2020-td,Yang2012-ze}. Because we wanted to compare our
inferences (based on intergenic sites) to inferences from previous work (based
on whole-genome data), we computed the cSFS for both intergenic and all sites
genome-wide. Sites with no calls in the Vindija Neanderthal were excluded from
this analysis. \aprcomment{Figures XX.}

@@

For the early history, we tested model parameterizations that cover many of the
proposed scenarios of population structure, size changes, and/or archaic
admixture. The simplest model, in terms of number of parameters, was a
single-origin expansion of modern humans, with no structure in the stem and no
archaic admixture aside from the Neanderthal admixture in Eurasian populations
following the out-of-Africa migration. This model allowed for a population size
change in the stem of modern humans between the ancestral split of the
human-Neanderthal lineages and the more recent split of branches leading to
Southern and West/Eastern African populations.

To include population structure in early human history, we considered multiple
parameterizations of models that allowed more than a single stem population. In
general, stem populations were allowed to vary in their sizes, split times, and
migration rates, with parameterizations flexible enough to encompass proposed
scenarios of either archaic admixture or population structure, both connected
by gene flow or with isolation between stems.

In one parameterization of early structure, which we refer to as a “continuous
migration” model, a secondary stem population (stem 2) split from the primary
stem (stem 1) that leads to modern humans. Stem 1 contributed to present-day
populations via a series of population splits similar to the single-origin
model, while stem 2 contributed through continuous symmetric migrations with
contemporaneous populations. The symmetrical migration rates could differ
across population pairs and over different epochs. This continuous migration
was allowed until stem 2 disappeared, which occurred as recently as 5kya. We
tested models that both allowed or disallowed migration between stems, i.e.,
before stem 1 split into S/E/W African populations.

In another parameterization of early structure, a secondary stem population
(stem 2) contributed ancestry to present-day populations via a series of
instantaneous admixture events (i.e., pulse admixture or “merger” events) to
lineages leading to sampled present-day African populations. Merger events were
allowed to occur in one or more of the Nama, Mende, and Gumuz branches, as well
as the branch of East and West Africans prior to their split. Those admixture
events were allowed to occur at any time along those branches, and with any
proportions, and stem 2 was allowed to split from the primary stem at any time
before subsequent divergences (and either before or after the split of the
Neanderthal branch). We tested models that allowed migration between the early
stems, before subsequent splits and admixture events, as well as models that
were restricted to isolation between stem branches. Depending on the specific
parameters, such models encompass commonly-considered ghost archaic admixture
scenarios (e.g., if a long-isolated lineage more recently contributes a
minority of ancestry to one or more populations), as well as relatively simple
fragmentation-coalescence scenarios.

Based on the geographical locations of present-day populations, we tentatively
labeled ancestral branches using a parsimony in migration, referring to South,
East, and West African branches (S/E/W AFR, Figure X). We do not know the
geographical location of these ancestral populations (nor even if they
correspond to populations with a well-defined geographic range), and these
labels should be considered as tentative. However, we found it useful to name
branches in reference to where in Africa their descendants are found.
\sgcomment{Please double-check this paragraph! TODO Brenna?}

@@

moments-LD uses a composite likelihood approach to simultaneously fit relative
pairwise diversity and LD statistics over a range of recombination distances.
Likelihoods were computed independently for pairwise diversity and each
recombination distance bin using a multivariate Guassian likelihood function as
described in Section \ref{sec:computing-stats}. These were multiplied across
bins and the single set of heterozygosity statistics following the approach
detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
rounds of optimization, alternating between the \emph{optimize\_log\_fim},
\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
parameter space and hone in on the best fit parameters. Initial guesses for
parameters were chosen from demographically plausible starting points and then
perturbed to explore space, using gradient descent (on the log of the
parameters). The best fits from these initial rounds of optimization were then
chosen as the starting points for optimization using the Powell and/or the
L-BFGS-B methods (as implemented in the SciPy optimization package
\cite{Virtanen2020-kr}). This process was repeated with alternating
optimization methods until the best fit parameter set converged consistently.

@@

The bootstrap replicates that were used to compute the variance-covariance
structure of the observed statistics within bins were also used to build 500
bootstrap replicates of the data by resampling with replacement. For the best
fit parameters, we computed confidence intervals using the Godambe Information
approach, which corrects composite likelihood estimates of confidence intervals
to account for nonindependence in the data, including linkages between loci and
nonindependence of recombination bins \cite{Coffman2016-yq}.

@@

\begin{figure}[ht]
\begin{center}
    \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
    \caption{\textbf{supp}.}
    \label{fig:supp-possible-models}
\end{center}
\end{figure}
